---
title: "Notes on likelihood"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
par(bg = 'black', fg = 'white', col.axis = 'white', col.lab = 'white', 
    col.main = 'white')
```

Here we document how to work with likelihoods.

First let's read in the data we generated from the multi-species competitive model.

```{r sad-data}
x <- readRDS('lvMultiSim.Rds')
plot(sort(x, decreasing = TRUE), log = 'y', 
     xlab = 'Species rank', ylab = 'Abundance')
```

We're going to fit the log-series distribution to the data. Here we make a *function* to compute the log-series.


```{r logseries-eq}
lseries <- function(b, n) {
    1 / log(1 / (1 - exp(-b))) * exp(-b * n) / n
}

hist(x, probability = TRUE)

points(1:50, lseries(0.1, 1:50), col = 'red', type = 'b')
points(1:50, lseries(0.5, 1:50), col = 'orange', type = 'b')
points(1:50, lseries(0.01, 1:50), col = 'blue', type = 'b')
```


A likelihood is simply the probability of observing the data.  Because these probabilities are typically very small, we most often work in log transformed probability, or *log likelihood*.

```{r likelihood-intro}
sum(log(lseries(0.001, x)))
sum(log(lseries(0.01, x)))
sum(log(lseries(0.5, x)))
```


```{r ll-curve, echo = FALSE}
bb <- seq(0.001, 0.1, length.out = 50)
ll <- sapply(bb, function(b) sum(log(lseries(b, x))))

pdf('fig_ll.pdf', width = 4, height = 4)
par(bg = 'black', fg = 'white', col.axis = 'white', col.lab = 'white')
plot(bb, ll, xlab = 'b parameter', ylab = 'log likelihood')
dev.off()
```


Fitting the model to the data precisely means finding the parameter value that maximizes the likelihood function.

```{r maxloglik}
llLSeries <- function(b, n) {
    sum(log(lseries(b, n)))
}


optimize(llLSeries, interval = c(0, 10), n = x, maximum = TRUE)
```

We can did all that likelihood by hand, but we can also use the pika package

```{r pika}
# devtools::install_github('ajrominger/pika')

library(pika)
s <- sad(x, 'lseries')
s
logLik(s)
```

The pika package let's us do nicer plots, and calculate a z-statistic, which is a goodness of fit measure.

```{r pika-fun}
plot(s, ptype = 'rad')
logLikZ(s)

pchisq(3.33, df = 1, lower.tail = FALSE)
```


```{r llz, eval = FALSE, echo = FALSE}
pdf('fig_lseriesSAD.pdf', width = 4, height = 4)
par(bg = 'black', fg = 'white', col.axis = 'white', col.lab = 'white')
plot(s, ptype = 'rad', thr.col = 'magenta')
dev.off()

pdf('fig_lseriesPerf.pdf', width = 4, height = 4)
par(bg = 'black', fg = 'white', col.axis = 'white', col.lab = 'white')
plot(sad(rlseries(s$nobs, s$MLE), 'lseries'), ptype = 'rad', thr.col = 'white')
dev.off()


llDist <- replicate(1000, sad(rlseries(s$nobs, s$MLE), 'lseries')$ll)
pdf('fig_llDist.pdf', width = 6, height = 4)
par(bg = 'black', fg = 'white', col.axis = 'white', col.lab = 'white')
curve(dnorm(x, mean = mean(llDist), sd = sd(llDist)), 
      from = -230, to = -130, yaxt = 'n', 
      xlab = 'log likelihood', ylab = '', lwd = 3)
dev.off()
```

Model comparison necessitates that we penalize models for their complexity (i.e. number of parameters). We do this with AIC

```{r model-comp}
ls <- sad(x, 'lseries')
ln <- sad(x, 'plnorm')
nb <- sad(x, 'tnegb')

AIC(ls)
AIC(ln)
AIC(nb)

plot(ls, ptype = 'rad', log = 'y', main = 'logseries')
plot(ln, ptype = 'rad', log = 'y', main = 'lognormal')
plot(nb, ptype = 'rad', log = 'y', main = 'negbinom')

logLikZ(ls)
logLikZ(nb)
```

Let's see if output from a neutral model looks any different

```{r neutral}
library(roleR)

p <- untbParams(1000, 100000, 200, 0.01, 0.1, 'oceanic_island', 
                50000, 50000)
neutMod <- roleModel(p)
neutMod <- iterModel(neutMod)
neutModFin <- getFinalState(neutMod)
y <- getSumStats(neutModFin, list(abund = rawAbundance))
y <- y$abund$abund
y <- y[y > 0]

untbNB <- sad(y, 'tnegb')
plot(untbNB, ptype = 'rad', log = 'y')
logLikZ(untbNB)
```
